\textbf{SPECFEM3D}\\ \textbf{Cartesian}\\ \textbf{User Manual}

### 2023-04-23 doc/USER_MANUAL/13_changing_the_model.tex

\noindent
Each data record line provides the velocity model values in a format like:
{\small
\begin{verbatim}
# data record format: purely isotropic model
x y z   vp    vs    rho
..
\end{verbatim}
}
\noindent
or
{\small
\begin{verbatim}
# data record format: anelastic isotropic model
x y z   vp    vs    rho Qp  Qs
..
\end{verbatim}
}
\noindent
where  \texttt{x}, \texttt{y}, \texttt{z} are the grid point position, \texttt{vp}, \texttt{vs} and \texttt{rho} the P- and S-wave speeds and density, and \texttt{Qp} and \texttt{Qs} the quality factors for P- and S-wave speeds. The quality factors are optional and can be omitted for purely elastic models (the tomography routine will recognize both formats). Internally, the quality factors will be converted to bulk and shear attenuation values,
$Q_{\kappa}$ and $Q_{\mu}$ respectively \citep{AnHa78}.
Note that Qmu is always equal to Qs, but Qkappa is in general not equal to Qp. To convert one to the other see doc/note\_on\_Qkappa\_versus\_Qp.pdf and utils/attenuation/conversion\_from\_Qkappa\_Qmu\_to\_Qp\_Qs\_from\_Dahlen\_Tromp\_959\_960.f90.
For simulations with attenuation, please note that the Vp- and Vs-velocities of your model are given for a reference
frequency. To change this reference frequency, you change the value
of \texttt{ATTENUATION\_f0\_REFERENCE} defined (in Hz) in input file \texttt{DATA/Par\_file}.

@@

The example in Figure~\ref{fig:tomography_file} is for an external tomography file with uniform spacing in the $x$, $y$, and $z$ directions. (Note that $\Delta x$, $\Delta y$, and $\Delta z$ need not be the same as in the example.) In that example, the  \texttt{nummaterial\_velocity\_file} is
%
\begin{verbatim}
2  -1 tomography elastic tomography_model.xyz 1
\end{verbatim}
%
and the file \texttt{tomography\_model.xyz} will need to reside in \texttt{DATA/}. All entries in the second column of \texttt{materials\_file} will be \texttt{-1}, which means that each element in the mesh will be interpolated according to the values in \texttt{tomography\_model.xyz}.

In some cases it may be desirable to have an external tomography model that is described in more than one file. For example, in cases like southern California, the length scale of variation in the structure of the wave speed model is much shorter in the sedimentary basin models within the upper 15~km. Therefore one might want to use an external tomography file that is sampled with, say, $\Delta x = 1000$~m, $\Delta y = 1000$~m, and $\Delta z = 250$~m in the uppermost 15~km, and then use $\Delta x = 2000$~m, $\Delta y = 2000$~m, and $\Delta z = 1000$~m below a depth of 15~km. If these intervals are chosen appropriately, then it will result in a pair of external tomography file that is much smaller than the alternative of having a single file with the fine discretization. In this case  \texttt{nummaterial\_velocity\_file} is
%
\begin{verbatim}
2  -1 tomography elastic file_above_15km.xyz 1
2  -2 tomography elastic file_below_15km.xyz 1
\end{verbatim}
%
and the files \texttt{file\_above\_15km.xyz} and \texttt{file\_below\_15km.xyz} will need to reside in \texttt{DATA/}.  All entries in the second column of \texttt{materials\_file} will need to be either \texttt{-1} or \texttt{-2}, depending on whether the element is above or below 15~km depth. In this sense, the construction of the mesh and the discretization of the external model will generally be done in tandem.

@@

Stanford Exploration Project (SEP) models consists of two files per variables, one header file (\texttt{.H})
and one binary file (\texttt{.H$@$}).
Among other information, the header file indicates, for each dimension,
the offset (\texttt{o1}, \texttt{o2} and \texttt{o3} fields),
the spatial step size (\texttt{d1}, \texttt{d2} and \texttt{d3} fields)
and the number of samples (\texttt{n1}, \texttt{n2} and \texttt{n3}) necessary to read
the linked binary file (\texttt{in} field).

SEP model reading occurs during database generation. It it the user responsibility
to ensure that proper interfaces between acoustic and elastic domains are set up
during the meshing phase. For more information, refer to
section~\ref{cha:Running-the-Mesher-Meshfem3D}.

In \texttt{DATA/Par\_file}, \texttt{\textbf{MODEL}} should be set to \texttt{\textbf{sep}},
and \texttt{\textbf{SEP\_MODEL\_DIRECTORY}} should point to wherever your sep model files reside.

Note that in order not to overload the parameter file, SEP header files
should be named \texttt{vp.H}, \texttt{vs.H} and \texttt{rho.H}. The \texttt{in} field
in these binary files can be any path as long as it is relative to
the previously set \texttt{SEP\_MODEL\_DIRECTORY}.
We also assume that any dimensional value inside SEP header files is given in meters and
that binary files are single precision floating point numbers stored in little endian format.
There is currently only support for \texttt{vp}, \texttt{vs} and \texttt{rho} variables.

>>>>

\noindent
Each data record line provides the velocity model values in a format like:
{\small
\begin{verbatim}
# data record format: purely isotropic model
x y z   vp    vs    rho
..
\end{verbatim}
}
\noindent
or
{\small
\begin{verbatim}
# data record format: anelastic isotropic model
x y z   vp    vs    rho Qp  Qs
..
\end{verbatim}
}
\noindent
where  \texttt{x}, \texttt{y}, \texttt{z} are the grid point position, \texttt{vp}, \texttt{vs} and \texttt{rho} the P- and S-wave speeds and density, and \texttt{Qp} and \texttt{Qs} the quality factors for P- and S-wave speeds. The quality factors are optional and can be omitted for purely elastic models (the tomography routine will recognize both formats). Internally, the quality factors will be converted to bulk and shear attenuation values,
$Q_{\kappa}$ and $Q_{\mu}$ respectively \citep{AnHa78}.
Note that Qmu is always equal to Qs, but Qkappa is in general not equal to Qp. To convert one to the other see doc/note\_on\_Qkappa\_versus\_Qp.pdf and utils/attenuation/conversion\_from\_Qkappa\_Qmu\_to\_Qp\_Qs\_from\_Dahlen\_Tromp\_959\_960.f90.
For simulations with attenuation, please note that the Vp- and Vs-velocities of your model are given for a reference
frequency. To change this reference frequency, you change the value
of \texttt{ATTENUATION\_f0\_REFERENCE} defined (in Hz) in input file \texttt{DATA/Par\_file}.\newline

@@

The example in Figure~\ref{fig:tomography_file} is for an external tomography file with uniform spacing in the $x$, $y$, and $z$ directions. (Note that $\Delta x$, $\Delta y$, and $\Delta z$ need not be the same as in the example.) In that example, the  \texttt{nummaterial\_velocity\_file} is
%
\begin{verbatim}
2  -1 tomography elastic tomography_model.xyz 1
\end{verbatim}
%
and the file \texttt{tomography\_model.xyz} will need to reside in \texttt{DATA/}. All entries in the second column of \texttt{materials\_file} will be \texttt{-1}, which means that each element in the mesh will be interpolated according to the values in \texttt{tomography\_model.xyz}.\newline

In some cases it may be desirable to have an external tomography model that is described in more than one file. For example, in cases like southern California, the length scale of variation in the structure of the wave speed model is much shorter in the sedimentary basin models within the upper 15~km. Therefore one might want to use an external tomography file that is sampled with, say, $\Delta x = 1000$~m, $\Delta y = 1000$~m, and $\Delta z = 250$~m in the uppermost 15~km, and then use $\Delta x = 2000$~m, $\Delta y = 2000$~m, and $\Delta z = 1000$~m below a depth of 15~km. If these intervals are chosen appropriately, then it will result in a pair of external tomography file that is much smaller than the alternative of having a single file with the fine discretization. In this case  \texttt{nummaterial\_velocity\_file} is
%
\begin{verbatim}
2  -1 tomography elastic file_above_15km.xyz 1
2  -2 tomography elastic file_below_15km.xyz 1
\end{verbatim}
%
and the files \texttt{file\_above\_15km.xyz} and \texttt{file\_below\_15km.xyz} will need to reside in \texttt{DATA/}.  All entries in the second column of \texttt{materials\_file} will need to be either \texttt{-1} or \texttt{-2}, depending on whether the element is above or below 15~km depth. In this sense, the construction of the mesh and the discretization of the external model will generally be done in tandem.\newline

@@

Stanford Exploration Project (SEP) models consists of two files per variables, one header file (\texttt{.H})
and one binary file (\texttt{.H$@$}).
Among other information, the header file indicates, for each dimension,
the offset (\texttt{o1}, \texttt{o2} and \texttt{o3} fields),
the spatial step size (\texttt{d1}, \texttt{d2} and \texttt{d3} fields)
and the number of samples (\texttt{n1}, \texttt{n2} and \texttt{n3}) necessary to read
the linked binary file (\texttt{in} field).\newline

SEP model reading occurs during database generation. It it the user responsibility
to ensure that proper interfaces between acoustic and elastic domains are set up
during the meshing phase. For more information, refer to
section~\ref{cha:Running-the-Mesher-Meshfem3D}.\newline

In \texttt{DATA/Par\_file}, \texttt{\textbf{MODEL}} should be set to \texttt{\textbf{sep}},
and \texttt{\textbf{SEP\_MODEL\_DIRECTORY}} should point to wherever your sep model files reside.\newline

Note that in order not to overload the parameter file, SEP header files
should be named \texttt{vp.H}, \texttt{vs.H} and \texttt{rho.H}. The \texttt{in} field
in these binary files can be any path as long as it is relative to
the previously set \texttt{SEP\_MODEL\_DIRECTORY}.
We also assume that any dimensional value inside SEP header files is given in meters and
that binary files are single precision floating point numbers stored in little endian format.
There is currently only support for \texttt{vp}, \texttt{vs} and \texttt{rho} variables.\newline

### 2023-04-23 doc/USER_MANUAL/15_informations_for_developers.tex

\noindent We also have simple 3D demo source codes that implement the SEM in a single, small program, in directory\\
utils/small\_SEM\_solvers\_in\_Fortran\_and\_C\_without\_MPI\_to\_learn of the specfem3d package.
They are useful to learn how the spectral-element method works, and how to write or modify a code to implement it.
Also useful to test new ideas by modifying these simple codes to run some tests.
We also have a similar, even simpler, demo source code for the 2D case in directory\\
utils/small\_SEM\_solver\_in\_Fortran\_without\_MPI\_to\_learn of the specfem2d package.\\

>>>>

\noindent
We also have simple 3D demo source codes that implement the SEM in a single, small program, in directory\newline
\texttt{utils/small\_SEM\_solvers\_in\_Fortran\_and\_C\_without\_MPI\_to\_learn} of the specfem3d package.
They are useful to learn how the spectral-element method works, and how to write or modify a code to implement it.
Also useful to test new ideas by modifying these simple codes to run some tests.
We also have a similar, even simpler, demo source code for the 2D case in directory\newline
\texttt{utils/small\_SEM\_solver\_in\_Fortran\_without\_MPI\_to\_learn} of the specfem2d package.\newline

### 2023-04-23 doc/USER_MANUAL/A_reference_frame.tex

The code uses the following convention for the Cartesian reference
frame:
\begin{itemize}
\item the $x$ axis points East
\item the $y$ axis points North
\item the $z$ axis points up
\end{itemize}
Note that this convention is different from both the \citet{AkRi80}
convention and the Harvard Centroid-Moment Tensor (CMT) convention.
The Aki \& Richards convention is
\begin{itemize}
\item the $x$ axis points North
\item the $y$ axis points East
\item the $z$ axis points down
\end{itemize}
and the Harvard CMT convention is
\begin{itemize}
\item the $x$ axis points South
\item the $y$ axis points East
\item the $z$ axis points up
\end{itemize}

>>>>

The code uses the following convention for the Cartesian reference
frame:
\begin{itemize}
\item the $x$ axis points East
\item the $y$ axis points North
\item the $z$ axis points up
\end{itemize}
Note that this convention is different from both the \citet{AkRi80}
convention and the Global/Harvard Centroid-Moment Tensor (CMT) convention.
The Aki \& Richards convention is
\begin{itemize}
\item the $x$ axis points North
\item the $y$ axis points East
\item the $z$ axis points down
\end{itemize}
and the Global/Harvard CMT convention is
\begin{itemize}
\item the $x$ axis points South
\item the $y$ axis points East
\item the $z$ axis points up
\end{itemize}

### 2023-04-23 doc/USER_MANUAL/B_channel_codes.tex
Path updates for parameters moved from constants.h to Par_file and vice versa, and other minor corrections.

Seismic networks, such as the Global Seismographic Network (GSN),
generally involve various types of instruments with different bandwidths,
sampling properties and component configurations. There are standards
to name channel codes depending on instrument properties. IRIS (\url{http://www.iris.edu})
uses SEED format for channel codes, which are represented by three
letters, such as \texttt{LHN}, \texttt{BHZ}, etc. In older versions
of the SPECFEM3D Cartesian package, a common format was used for the
channel codes of all seismograms, which was \texttt{BHE/BHN/BHZ} for
three components. To avoid confusion when comparison are made to observed
data, we are now using the FDSN convention (\url{http://www.fdsn.org/})
for SEM seismograms. In the following, we give a brief explanation
of the FDSN convention used by IRIS, and how it is adopted in SEM
seismograms. Please visit \url{http://www.iris.edu/manuals/SEED_appA.htm}
for further information.\\

\noindent \texttt{\textbf{Band code:}} The first letter in the channel
code denotes the band code of seismograms, which depends on the response
band and the sampling rate of instruments. The list of band codes
used by IRIS is shown in Figure \ref{fig:IRIS_band_codes}. The sampling
rate of SEM synthetics is controlled by the resolution of simulations
rather than instrument properties. However, for consistency, we follow
the FDSN convention for SEM seismograms governed by their sampling
rate. For SEM synthetics, we consider band codes for which $dt\leq1$
s. IRIS also considers the response band of instruments. For instance,
short-period and broad-band seismograms with the same sampling rate
correspond to different band codes, such as S and B, respectively.
In such cases, we consider SEM seismograms as broad band, ignoring
the corner period ($\geq10$ s) of the response band of instruments
(note that at these resolutions, the minimum period in the SEM synthetics
will be less than $10$ s). Accordingly, when you run a simulation
the band code will be chosen depending on the resolution of the synthetics,
and channel codes of SEM seismograms will start with either \texttt{L},
\texttt{M}, \texttt{B}, \texttt{H}, \texttt{C} or \texttt{F}, shown
by red color in the figure.\\

@@

\noindent \texttt{\textbf{Instrument code:}} The second letter in
the channel code corresponds to instrument codes, which specify the
family to which the sensor belongs. For instance, \texttt{H} and \texttt{L}
are used for high-gain and low-gain seismometers, respectively. The
instrument code of SEM seismograms will always be \texttt{X}, as assigned
by FDSN for synthetic seismograms. \\

\noindent \texttt{\textbf{Orientation code:}} The third letter in
channel codes is an orientation code, which generally describes the
physical configuration of the components of instrument packages. SPECFEM3D
Cartesian uses the traditional orientation code \texttt{E/N/Z} (East-West,
North-South, Vertical) for three components when a UTM projection
is used. If the UTM conversion is suppressed, i.e. the flag \texttt{SUPPRESS\_UTM\_PROJECTION}
is set to \texttt{.true.}, then the three components are labelled
\texttt{X/Y/Z} according to the Cartesian reference frame. \\

\noindent \texttt{\textbf{EXAMPLE:}} The sampling rate is given by
\texttt{DT} in the main parameter file \texttt{Par\_file} located
in the \texttt{DATA/} subdirectory. Depending on the resolution of
your simulations, if you choose a sampling rate greater than $0.01$
s and less than $1$ s, a seismogram recording displacements on the
vertical component of a station \texttt{ASBS} (network \texttt{AZ})
will be named \texttt{AZ.ASBS.MXZ.semd.sac}, whereas it will be \texttt{AZ.ASBS.BXZ.semd.sac},
if the sampling rate is greater than 0.0125 and less equal to 0.1
s.

>>>>

Seismic networks, such as the Global Seismographic Network (GSN),
generally involve various types of instruments with different bandwidths,
sampling properties and component configurations. There are standards
to name channel codes depending on instrument properties. IRIS (\url{http://www.iris.edu})
uses SEED format for channel codes, which are represented by three
letters, such as \texttt{LHN}, \texttt{BHZ}, etc. In older versions
of the SPECFEM3D Cartesian package, a common format was used for the
channel codes of all seismograms, which was \texttt{BHE/BHN/BHZ} for
three components. To avoid confusion when comparison are made to observed
data, we are now using the FDSN convention (\url{http://www.fdsn.org/})
for SEM seismograms. In the following, we give a brief explanation
of the FDSN convention used by IRIS, and how it is adopted in SEM
seismograms. Please visit \url{http://www.iris.edu/manuals/SEED_appA.htm}
for further information.\newline

\noindent \texttt{\textbf{Band code:}} The first letter in the channel
code denotes the band code of seismograms, which depends on the response
band and the sampling rate of instruments. The list of band codes
used by IRIS is shown in Figure \ref{fig:IRIS_band_codes}. The sampling
rate of SEM synthetics is controlled by the resolution of simulations
rather than instrument properties. However, for consistency, we follow
the FDSN convention for SEM seismograms governed by their sampling
rate. For SEM synthetics, we consider band codes for which $dt\leq1$
s. IRIS also considers the response band of instruments. For instance,
short-period and broad-band seismograms with the same sampling rate
correspond to different band codes, such as S and B, respectively.
In such cases, we consider SEM seismograms as broad band, ignoring
the corner period ($\geq10$ s) of the response band of instruments
(note that at these resolutions, the minimum period in the SEM synthetics
will be less than $10$ s). Accordingly, when you run a simulation
the band code will be chosen depending on the resolution of the synthetics,
and channel codes of SEM seismograms will start with either \texttt{L},
\texttt{M}, \texttt{B}, \texttt{H}, \texttt{C} or \texttt{F}, shown
by red color in the figure.\newline

@@

\noindent \texttt{\textbf{Instrument code:}} The second letter in
the channel code corresponds to instrument codes, which specify the
family to which the sensor belongs. For instance, \texttt{H} and \texttt{L}
are used for high-gain and low-gain seismometers, respectively. The
instrument code of SEM seismograms will always be \texttt{X}, as assigned
by FDSN for synthetic seismograms. \newline

\noindent \texttt{\textbf{Orientation code:}} The third letter in
channel codes is an orientation code, which generally describes the
physical configuration of the components of instrument packages. SPECFEM3D
Cartesian uses the traditional orientation code \texttt{E/N/Z} (East-West,
North-South, Vertical) for three components when a UTM projection
is used. If the UTM conversion is suppressed, i.e. the flag \texttt{SUPPRESS\_UTM\_PROJECTION}
is set to \texttt{.true.}, then the three components are labelled
\texttt{X/Y/Z} according to the Cartesian reference frame. \newline

\noindent \texttt{\textbf{EXAMPLE:}} The sampling rate is given by
\texttt{DT} in the main parameter file \texttt{DATA/Par\_file}. Depending on the resolution of
your simulations, if you choose a sampling rate greater than $0.01$
s and less than $1$ s, a seismogram recording displacements on the
vertical component of a station \texttt{ASBS} (network \texttt{AZ})
will be named \texttt{AZ.ASBS.MXZ.semd.sac}, whereas it will be \texttt{AZ.ASBS.BXZ.semd.sac},
if the sampling rate is greater than 0.0125 and less equal to 0.1
s.

### 2023-04-23 doc/USER_MANUAL/authors.tex

\section*{https://specfem3d.readthedocs.io/en/latest/authors/}

\noindent The SPECFEM3D package was first developed by Dimitri
Komatitsch and Jean-Pierre Vilotte at Institut de Physique du Globe
(IPGP) in Paris, France from 1995 to 1997 and then by Dimitri Komatitsch
and Jeroen Tromp at Harvard University and Caltech, USA, starting in 1998.
The story started on April 4, 1995, when Prof. Yvon Maday from CNRS and University of Paris, France, gave a lecture to
Dimitri Komatitsch and Jean-Pierre Vilotte at IPG about the nice properties of the Legendre spectral-element method with diagonal mass matrix that he had used for
other equations. We are deeply indebted and thankful to him for that.
That followed a visit by Dimitri Komatitsch to OGS (Istituto Nazionale di Oceanografia e di Geofisica Sperimentale) in Trieste, Italy, in February 1995
to meet with G\'eza Seriani and Enrico Priolo, who introduced him to their 2D Chebyshev version of the spectral-element method with a non-diagonal mass matrix.
We are deeply indebted and thankful to them for that.

%%%%%%%% please do NOT merge the lines of the paragraph below, it makes it difficult to insert new names in alphabetical order; leave a single name per line
%%%%%%%% please do NOT merge the lines of the paragraph below, it makes it difficult to insert new names in alphabetical order; leave a single name per line
%%%%%%%% please do NOT merge the lines of the paragraph below, it makes it difficult to insert new names in alphabetical order; leave a single name per line
Nikhil Vaidyar

The cover graphic of the manual was created by Santiago Lombeyda from
Caltech's Center for Advanced Computing Research (CACR), USA, with free satellite clipart pictures
from \url{http://www.4vector.com} and \url{http://www.clker.com} added to it.\\

The code is released open-source under the GNU version 3 license, see the license at the end of this manual.\\

>>>>

\section*{Authors}

\noindent The SPECFEM3D package was first developed by Dimitri
Komatitsch and Jean-Pierre Vilotte at Institut de Physique du Globe
(IPGP) in Paris, France from 1995 to 1997 and then by Dimitri Komatitsch
and Jeroen Tromp at Harvard University and Caltech, USA, starting in 1998.
The story started on April 4, 1995, when Prof. Yvon Maday from CNRS and University of Paris, France, gave a lecture to
Dimitri Komatitsch and Jean-Pierre Vilotte at IPG about the nice properties of the Legendre spectral-element method with diagonal mass matrix that he had used for
other equations. We are deeply indebted and thankful to him for that.
That followed a visit by Dimitri Komatitsch to OGS (Istituto Nazionale di Oceanografia e di Geofisica Sperimentale) in Trieste, Italy, in February 1995
to meet with G\'eza Seriani and Enrico Priolo, who introduced him to their 2D Chebyshev version of the spectral-element method with a non-diagonal mass matrix.
We are deeply indebted and thankful to them for that.\newline

%%%%%%%% please do NOT merge the lines of the paragraph below, it makes it difficult to insert new names in alphabetical order; leave a single name per line
%%%%%%%% please do NOT merge the lines of the paragraph below, it makes it difficult to insert new names in alphabetical order; leave a single name per line
%%%%%%%% please do NOT merge the lines of the paragraph below, it makes it difficult to insert new names in alphabetical order; leave a single name per line
Since then it has been developed and maintained by a development team: in alphabetical order,
Michael Afanasiev,
Rafael Almada,
Jean-Paul (Pablo) Ampuero,
\'Etienne Bachmann,
Kangchen Bai,
Piero Basini,
Stephen Beller,
Jordan Bishop,
C\'eline Blitz,
Alexis Bottero,
Ebru Bozda\u{g},
Emanuele Casarotti,
Joseph Charles,
Min Chen,
Paul Cristini,
Cl\'ement Durochat,
Percy Galvez,
Rene Gassmoeller,
Hom Nath Gharti,
Dominik G\"oddeke,
Leopold Grinberg,
Aakash Gupta,
Vala Hj\"orleifsd\'ottir,
Foivos Karakostas,
Sue Kientz,
Dimitri Komatitsch,
Jes\'us Labarta,
Nicolas Le Goff,
Pieyre Le Loher,
Matthieu Lefebvre,
Qinya Liu,
Youshan Liu,
David Luet,
Yang Luo,
Alessia Maggi,
Federica Magnoni,
Roland Martin,
Ren\'e Matzen,
Geordie McBain,
Dennis McRitchie,
Matthias Meschede,
Peter Messmer,
David Mich\'ea,
Ryan Modrak,
Vadim Monteiller,
Surendra Nadh Somala,
Masaru Nagaso,
Tarje Nissen-Meyer,
Elif Oral,
Daniel Peter,
Kevin Pouget,
Max Rietmann,
Elliott Sales de Andrade,
Brian Savage,
Bernhard Schuberth,
Anne Sieminski,
James Smith,
Leif Strand,
Carl Tape,
Jeroen Tromp,
Eduardo Valero Cano,
Brice Videau,
Jean-Pierre Vilotte,
Huihui Weng,
Zhinan Xie,
Chang-Hua Zhang,
Hejun Zhu.\newline

The cover graphic of the manual was created by Santiago Lombeyda from
Caltech's Center for Advanced Computing Research (CACR), USA, with free satellite clipart pictures
from \url{http://www.4vector.com} and \url{http://www.clker.com} added to it.\newline

The code is released open-source under the GNU version 3 license, see the license at the end of this manual.\newline
